Bayesian regression facilitates quantitative modelling of cell metabolism

### 2023-10-31 paper/main.qmd
paper
Update paper

Importantly, this form of analysis takes into account all modelled sources of
uncertainty, including uncertainty about the measured values of the flux in each
dataset. Our result shows that Maud could be used in this realistic case not
only to provide a user with an explanation for an observed difference in fluxes,
but also a reasonable judgement as to the explanation's robustness.

@@

Maud’s prior model includes unknown parameters corresponding to quantities
in the kinetic model that are assumed to be unknown, other than steady state
metabolite concentrations and fluxes, which are derived from the values of
other parameters by solving the steady state problem. See table (REFERENCE) in
this paper's supporting information for a description of all these parameters
and their dimensions. Note that some quantities in Maud’s kinetic model are
not treated as parameters: for example, temperatures, compartment volumes and
the formation energy of water. Maud treats these quantities as if they were
known precisely: they can be configured by the user or default values can be
used. Although in practice there can be considerable uncertainty regarding
these quantities, we chose to disregard this uncertainty in the interest of
simplicity.

@@

Maud's parameters are connected with measurable quantities via the steady state
equation \eqref{eq-steady}. Posterior sampling using adaptive Hamiltonian Monte
Carlo requires repeatedly solving equation \eqref{eq-steady} and finding the
gradients of its solution with respect to all model parameters. This must be
done numerically as analytic solutions are not available for realistic kinetic
models.

@@

To solve equation \eqref{eq-steady} and find its gradients, Maud uses
a hybrid method involving two numerical solvers from the SUNDIALS
suite [@serbanCVODESSensitivityEnabledODE2005]: CVODES and IDAS via
their interface from Stan. The hybrid method follows that proposed by
@margossianComputingSteadyStates2018, and involves numerically evolving the ODE
system for a short period of time, then using the difference between the evolved
and starting concentrations as the target for a numerical algebra solver.

@@

- Description of Maud's input format
- Description of Maud's kinetic model including rate equations as well as
parameters and their dimensions.
- Details of the procedure used to generate our case study results, including
generating the artificial case study datasets, specifying priors and carrying
out computation.
- Specification of the prior distribution used for the case studies.

>>>>

Importantly, this form of analysis takes into account all modelled sources of
uncertainty, including uncertainty about the measured values of the flux in
each dataset. Our result shows that Maud could be used in this realistic case
not only to explain an observed difference in fluxes, but also to provide a
reasonable judgement as to the explanation's robustness.

@@

Maud’s prior model includes unknown parameters corresponding to quantities
in the kinetic model that are assumed to be unknown, other than steady state
metabolite concentrations and fluxes, which are derived from the values of
other parameters by solving the steady state problem. See table 1 in
this paper's supporting information for a description of all these parameters
and their dimensions. Note that some quantities in Maud’s kinetic model are
not treated as parameters: for example, temperatures, compartment volumes and
the formation energy of water. Maud treats these quantities as if they were
known precisely: they can be configured by the user or default values can be
used. Although in practice there can be considerable uncertainty regarding
these quantities, we chose to disregard this uncertainty in the interest of
simplicity.

@@

Maud's parameters are connected with measurable quantities via the steady state
equation \eqref{eq-3}. Posterior sampling using adaptive Hamiltonian Monte
Carlo requires repeatedly solving equation \eqref{eq-3} and finding the
gradients of its solution with respect to all model parameters. This must be
done numerically as analytic solutions are not available for realistic kinetic
models.

@@

To solve equation \eqref{eq-3} and find its gradients, Maud uses
a hybrid method involving two numerical solvers from the SUNDIALS
suite [@serbanCVODESSensitivityEnabledODE2005]: CVODES and IDAS via
their interface from Stan. The hybrid method follows that proposed by
@margossianComputingSteadyStates2018, and involves numerically evolving the ODE
system for a short period of time, then using the difference between the evolved
and starting concentrations as the target for a numerical algebra solver.

@@

- Description of Maud's input format.
- Description of Maud's kinetic model including rate equations as well as
parameters and their dimensions.
- Details of the procedure used to generate our case study results, including
generating the artificial case study datasets, specifying priors and carrying
out computation.
- Specification of the prior distribution used for the case studies.

### 2023-10-31 paper/main.tex
Update paper

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Bayesian regression facilitates quantitative modelling of cell metabolism},
  pdfauthor={Teddy Groves; Nicholas Luke Cowie; Lars Keld Nielsen},
  pdfkeywords={Bayesian inference, kinetic models of cell metabolism},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\keywords{Bayesian inference, kinetic models of cell metabolism}

\title[]{Bayesian regression facilitates quantitative modelling of cell
metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Our paper reviews the existing options in this
area, presents a case study illustrating how Maud can be used to analyse
a metabolic network and explains the biological, statistical and
computational design decisions underpinning Maud.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable]}{\end{tcolorbox}}\fi

@@

Maud's prior model includes unknown parameters corresponding to
quantities in the kinetic model that are assumed to be unknown, other
than steady state metabolite concentrations and fluxes, which are
derived from the values of other parameters by solving the steady state
problem. See table (REFERENCE) in this paper's supporting information
for a description of all these parameters and their dimensions. Note
that some quantities in Maud's kinetic model are not treated as
parameters: for example, temperatures, compartment volumes and the
formation energy of water. Maud treats these quantities as if they were
known precisely: they can be configured by the user or default values
can be used. Although in practice there can be considerable uncertainty
regarding these quantities, we chose to disregard this uncertainty in
the interest of simplicity.

@@

To solve equation \eqref{eq-steady} and find its gradients, Maud uses a
hybrid method involving two numerical solvers from the SUNDIALS suite
\citep{serbanCVODESSensitivityEnabledODE2005}: CVODES and IDAS via their
interface from Stan. The hybrid method follows that proposed by
\citet{margossianComputingSteadyStates2018}, and involves numerically
evolving the ODE system for a short period of time, then using the
difference between the evolved and starting concentrations as the target
for a numerical algebra solver.

@@

\begin{itemize}
\tightlist
\item
  Description of Maud's input format
\item
  Description of Maud's kinetic model including rate equations as well
  as parameters and their dimensions.
\item
  Details of the procedure used to generate our case study results,
  including generating the artificial case study datasets, specifying
  priors and carrying out computation.
\item
  Specification of the prior distribution used for the case studies.
\end{itemize}

>>>>

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Bayesian regression facilitates quantitative modelling of cell metabolism},
  pdfauthor={Teddy Groves; Nicholas Luke Cowie; Lars Keld Nielsen},
  pdfkeywords={Bayesian inference, kinetic models of cell metabolism,
multiomics integration, ordinary differential equations, regulatory
anlaysis},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\keywords{Bayesian inference, kinetic models of cell metabolism,
multiomics integration, ordinary differential equations, regulatory
anlaysis}

\title[]{Bayesian regression facilitates quantitative modelling of cell
metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Our paper reviews the existing options in this
area, presents a case study illustrating how Maud can be used to analyse
a metabolic network and explains the biological, statistical and
computational design decisions underpinning Maud.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, sharp corners, breakable, frame hidden]}{\end{tcolorbox}}\fi

@@

Maud's prior model includes unknown parameters corresponding to
quantities in the kinetic model that are assumed to be unknown, other
than steady state metabolite concentrations and fluxes, which are
derived from the values of other parameters by solving the steady state
problem. See table 1 in this paper's supporting information for a
description of all these parameters and their dimensions. Note that some
quantities in Maud's kinetic model are not treated as parameters: for
example, temperatures, compartment volumes and the formation energy of
water. Maud treats these quantities as if they were known precisely:
they can be configured by the user or default values can be used.
Although in practice there can be considerable uncertainty regarding
these quantities, we chose to disregard this uncertainty in the interest
of simplicity.

@@

To solve equation \eqref{eq-3} and find its gradients, Maud uses a
hybrid method involving two numerical solvers from the SUNDIALS suite
\citep{serbanCVODESSensitivityEnabledODE2005}: CVODES and IDAS via their
interface from Stan. The hybrid method follows that proposed by
\citet{margossianComputingSteadyStates2018}, and involves numerically
evolving the ODE system for a short period of time, then using the
difference between the evolved and starting concentrations as the target
for a numerical algebra solver.

@@

\begin{itemize}
\tightlist
\item
  Description of Maud's input format.
\item
  Description of Maud's kinetic model including rate equations as well
  as parameters and their dimensions.
\item
  Details of the procedure used to generate our case study results,
  including generating the artificial case study datasets, specifying
  priors and carrying out computation.
\item
  Specification of the prior distribution used for the case studies.
\end{itemize}
